Designing Types for R, Empirically

### 2020-05-16 paper/main.tex
scalara

\paragraph{NAs}
Our data supports the choice of making the presence of NAs explicit.  In
practice, programmers explicitly check for NAs in primitive vectors,
possibly sanitizing them if they are present.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
#' @type <x: dbl[], n: dbl[], conf.level: dbl, maxsteps: dbl,
#'         del: dbl, bayes: lgl, plot: lgl, ...> => class<data.frame>
binom.profile <- function(x, n, conf.level = 0.95, maxsteps = 50,
                            del = zmax / 5, bayes = TRUE, plot = FALSE, ...) {
   xn <- cbind(x = x, n = n)
   ok <- !is.na(xn[, 1]) & !is.na(xn[, 2])
   x <- xn[ok, "x"]
   n <- xn[ok, "n"]
\end{lstlisting}

The \code{binom} package is a popular R package for computing confidence
intervals, and the \code{binom.profile} function does so using the profile
likelihood.  This code snippet highlights a data sanitization pattern: the
programmer first binds the vectors into a matrix, with one column for each
of \code{x} and \code{n}, then finds rows where both columns are not \NA,
then extracts non-\NA values and stores them into \code{x} and \code{n}
respectively.We see from the type that arguments \code{x} and \code{n} are intended to be free of \NA values.
\contractr will check them for compliance, and warn programmers of function misuse.

\paragraph{Scalars}
The data also suggests that programmers often use scalars, and do
dimensionality checks on their data.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
#' @type <n: dbl, x: int[]> => class<matrix>
hankel.matrix <- function( n, x ) { 
  ### n = a positive integer value for the order of the Hankel matrix
  ### x = an order 2 * n + 1 vector of numeric values
    ...
    if ( n != trunc( n ) )  stop( "argument n ix not an integer" )
    if ( !is.vector( x ) )  stop( "argument x is not a vector" )
    m <- length( x )
    if ( m < n )  stop( "length of argument x is less than n" )
\end{lstlisting}

The \code{hankel.matrix} function takes two arguments.  The function has a
number of interesting checks: \code{n} undergoes an integer type check;
\code{x} undergoes a vector type check; and \code{n} is expected to be a
scalar, as if it were a vector, the code \code{m < n} would essentially
check \code{m < n[i]} for each element in \code{n}, returning a vector of
comparison results, and the conditional of an if must be a scalar else R
generates a warning. The type for this function indicates that \code{x} must be
a vector, though here the type for \code{n} is found to be \dbl---this is likely due
to numeric values defaulting to \dbl type in R, as e.g. \code{typeof(2)} is \dbl.

@@

Matrices represent another layer of complexity on top of vectorized
primitives.  Internally, matrices are simply vectors with class ``matrix''
and a \code{dims} attribute indicating matrix dimensions, and while not
codified in the language semantics, many internal functions will coerce
vectors to matrices automatically.  Relying on automatic coercion is unwise,
and many functions check to ensure that their arguments are matrices.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
#' @type <x: class<matrix>, w: ? ^dbl[], rows: null, cols: null, na.rm: lgl, ...> => ^dbl[]
rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
                             na.rm = FALSE, ...) {
  if (!is.matrix(x)) 
    .Defunct(msg = sprintf("Argument 'x' should be a matrix.)
\end{lstlisting}

The \code{rowWeightedMeans} function calculates the weighted means of rows
or columns of a matrix \code{x}.  This function exhibits good practice in
type checking arguments, rightly producing a message to the user indicating
that an explicit matrix should be passed. We captured this with our class types, 
and indeed \typetracer inferred a good type for this function, allowing \contractr to 
perform the check present in this code automatically, notifying the users passing ill-typed arguments.

@@

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL, na.rm = FALSE) {
    ...
    n <- ncol(x)
    if (length(w) != n) 
       stop("The length of 'w' does not match columns in 'x': ")
\end{lstlisting}

This code is from later in the same \code{rowWeightedMeans} function
discussed previously.  Here, the length of the weights vector \code{w} is
being checked for equality with the number of columns in the matrix
\code{x}.  Specifying this relationship as a type amounts to {\it dependent
  typing}, which we believe too complex to retrofit onto a dynamic data
science language such as R.  Ultimately, we decided to leave these kinds of
checks up to the programmer.

@@

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
#' @type <x: class<aov, lm> | class<lm>> => dbl;
agricolae::cv.model <- function(x) {
    suma2 <- sum(x$residual^2)
    gl <- x$df.residual
    promedio <- mean(x$fitted.values)
    return(sqrt(suma2/gl)*100/promedio)
}
data(sweetpotato)
model<-aov(yield~virus, data=sweetpotato)
cv.model(model)
\end{lstlisting}

@@

%
\paragraph{Data Frames}

>>>>

\paragraph{NAs}
Our data supports making the presence of {\NA}s explicit. Only 1,058 (of
0.9\%) of arguments are marked as possibly having {\NA}s, thus the
overwhelming majority of types appear to be {\NA}-free.  In practice,
programmers check for them and sanitize them if they are present.  Consider
the \code{binom} package for computing confidence intervals and its
\code{binom.profile} function.  This attached code snippet highlights a data
sanitization pattern: the programmer first binds the vectors into a matrix,
then finds rows where both columns are not \NA, extracts non-\NA values and
stores them into \code{x} and \code{n} respectively.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  binom.profile <- function(x, n, conf.level=0.9, maxsteps=50,...) {
    xn <- cbind(x = x, n = n)
    ok <- !is.na(xn[, 1]) & !is.na(xn[, 2])
    x <- xn[ok, "x"]
    n <- xn[ok, "n"]
\end{lstlisting}

\paragraph{Scalars}
The data also suggests that programmers often use scalars, and do
dimensionality checks on their data.  In our data 31,533 (or 29\%) of the
arguments are scalar types. While not completely surprising, this is a
rather large number. Consider the \code{hankel.matrix} function, it takes
two arguments and checks that \code{n} is \int, that \code{x} is a vector,
and also, indirectly, that \code{n} is a a scalar (this comes from the fact
that it is used in the guard of a conditional which fails if \code{n} is
not a scalar).

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
 hankel.matrix <- function( n, x ) { 
  ### n = a positive integer value for the order of the Hankel matrix
  ### x = an order 2 * n + 1 vector of numeric values
    if ( n != trunc( n ) )  stop( "argument n ix not an integer" )
    if ( !is.vector( x ) )  stop( "argument x is not a vector" )
    m <- length( x )
    if ( m < n )  stop( "length of argument x is less than n" )
\end{lstlisting}

@@

Matrices represent another layer of complexity on top of vectorized
primitives.  Internally, matrices are simply vectors with class ``matrix''
and a \code{dims} attribute indicating matrix dimensions, and while not
codified in the language semantics, many internal functions will coerce
vectors to matrices automatically.  Relying on automatic coercion is unwise,
and many functions check to ensure that their arguments are matrices.
The \code{rowWeightedMeans} function calculates the weighted means of rows
or columns of a matrix \code{x}.  This function exhibits good practice in
type checking arguments, rightly producing a message to the user indicating
that an explicit matrix should be passed. We captured this with our class
types, and indeed \typetracer inferred a good type for this function,
allowing \contractr to perform the check present in this code automatically,
notifying the users passing ill-typed arguments.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  rowWeightedMeans <- function(x, w=NULL, rows=NULL, cols=NULL, na.rm=FALSE, ...) {
    if (!is.matrix(x)) 
      .Defunct(msg = sprintf("Argument 'x' should be a matrix.)
\end{lstlisting}

@@

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL, na.rm = FALSE) {
    n <- ncol(x)
    if (length(w) != n) 
       stop("The length of 'w' does not match columns in 'x': ")
\end{lstlisting}

This code is from later in the same \code{rowWeightedMeans} function
pdiscussed previously.  Here, the length of the weights vector \code{w} is
being checked for equality with the number of columns in the matrix
\code{x}.  Specifying this relationship as a type amounts to {\it dependent
  typing}, which we believe too complex to retrofit onto a dynamic data
science language such as R.  Ultimately, we decided to leave these kinds of
checks up to the programmer.

@@

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  agricolae::cv.model <- function(x) {
    suma2 <- sum(x$residual^2)
    gl <- x$df.residual
    promedio <- mean(x$fitted.values)
    return(sqrt(suma2/gl)*100/promedio)
  }
  data(sweetpotato)
  model<-aov(yield~virus, data=sweetpotato)
  cv.model(model)
\end{lstlisting}

@@

\paragraph{Data Frames}
